Finally Tagless, Partially Evaluated

### 2007-07-27 tagless/tagless-final.tex
added write-up about state-passing CPS

We can modify the CBV CPS transformation to pass a piece of state along
with the continuation. This technique lets us support mutable state. We
can also add mutable references to the object language using mutable
references of the metalanguage.  The accompanying code illustrates these
extensions.  No new techniques were needed in the implementation.

>>>>

We can modify the CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state and implement an extended object language, with operations
|deref ()| to obtain the current state, |set v| to replace the current
state with |v| returning the previous value of the state, and the
let-form |let x = e1 in e2|. The latter enforces the order of
evaluation: |e1| will always be evaluated before |e2| even if the
expression |e2| does not need the value of |x|. The more familiar
imperative sequencing form |begin e1; e2 end| is expressible in our
object language as |let _ = e1 in e2|.  In HOAS, the let-form actually
has the form |lapp e1 (fun x -> e2)|, which looks like the functional
application with its two arguments swapped.  The extended object
language corresponds to the following signature, which
includes all of the original Symantics and adds imperative features.

\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states	(* static version of the state *)
  val lapp : ('c,'sa,'da) repr -> 
             (('c,'sa,'da) repr -> ('c,'sb,'db) repr) ->
             ('c,'sb,'db) repr
  val deref : unit -> ('c,'c states,state) repr
  val set   : (('c,'c states,state) repr as 'x) -> 'x
end
\end{code}

We can write a simple test: 
\begin{code}
let v0 = !state in state := 2; v0 + !state
\end{code}
which, in our language, takes the form
\begin{code}
module EXSI_INT(S: SymSI with type state = int and type 'c states = int) = 
struct
  open S
  let test1 () = lapp (deref ()) (fun v0 -> 
                  lapp (set (int 2)) (fun _ ->
		   add v0 (deref ())))
end;;
\end{code}
The accompanying source code shows several more tests, including the
imperative power function that uses the state as the accumulator and
the test of a higher-order state.

The state-passing interpreter is the extension of the CBN CPS
interpreter RCN of\S\ref{S:CPS}:
\begin{code}
module RCPS (ST: sig 
  type state 
  type 'c states 
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end) = struct
  include ST
  type ('c, 'sv, 'dv) result = 'c states -> 'sv
  ... 
  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
  let lapp e2 e1 = 
    {ko = fun k -> e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
  let deref () = {ko = fun k s -> k s s}
  let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
end
\end{code}
The implementations of |int|, |app|, |lam|, etc. functions is \emph{identical}
to those of |RCN| and elided. New here is the extended type |repr|
that now includes the state, and the `constructors' of the imperative
forms of the object language: |lapp|, |deref|, and |set|. The
interpreter is still CBN and so evaluation of |app ef ea| does not
guarantee to evaluate |ea|. On the other hand, the implementation of
the let-form |lapp ea ef| does assure that |ea| will always be
evaluated. In the case of the first-order state such as |int| we 
instantiate the interpreter as
\begin{code}
module RCPSI = RCPS(struct 
  type state = int
  type 'c states = int
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end);;
\end{code}
It fully implements the |Symantics| signature and so can run the pure
code of the unextended object language: we can pass |RCPSI| as an
argument to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
the example |test1| from there. The main use for |RCPSI| is to run the 
extended object language:
\begin{code}
module EXPSI_INT = EXSI_INT(RCPSI)
let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100 (* 102 *)
\end{code}

We can also add mutable references to the object language using
mutable references of the metalanguage, as shown in the accompanying
code.  No new techniques were needed in the implementation.
